Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-09-07 paper-correlations.tex
Forgot to increment l
Fix a typo

The interface between phases must not change rapidly for the edge filter to
work correctly. Therefore, the algorithms for surface functions require input
images to be with an apropriate resolution. A criterion for goodness can be expressed with
$C_{\alpha}$ parameter. Let $\alpha \in [0, 1]$ and $f(x)$ be
(band-limited) input image ($x \in \mathbb{R}^2$ or $\mathbb{R}^3$). We define
$C_\alpha$ as follows:
\begin{align*}
  f_0(x) &= f(x) - \langle f(x) \rangle \\
  C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
    dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
\end{align*}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. We have previously
shown that criterion $C_{0.5} > 0.97$ holds then the image is suitable for
robust $F_{ss}$ and $F_{sv}$ evaluation \cite{samarin2023robust}.  Note that
function \code{lowfreq\_energy\_ratio} that computes $C_{0.5}$ parameter for any
input binary image is the part of \code{CorrelationFunctions.jl} package.

@@

Cluster function can be calculated using the same technique as in map approach,
but if the number of clusters is high, a naïve $O(n^2)$ algorithm may be
preferred to $O(MC \log M)$ full map algorithm. Such an approach is present below:
\begin{algorithmic}[1]
  \Procedure{c2}{$A, phase, periodic$}
    \State $A' \gets I^{(phase)}(A)$
    \State $CC \gets label\_components(A', periodic)$
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \ForAll{$slice \in slices(CC)$}
      \For{$i = 0,len-1$}
        \State $end \gets len-1$
        \If{$periodic$}
          \State $end \gets end + i$
        \EndIf
        \For{$j = i, end$}
          \If{$slice[i] = slice[j \mod len] \ne 0$}
            \Comment $0$ is a label for background (phase not equal to $phase$).
            \State $dst \gets |(j \mod len) - i|$
            \State $result[dst] \gets result[dst] + 1$
          \EndIf
        \EndFor
      \EndFor
    \EndFor
    \State \textbf{return} $result$ divided by number of trials
  \EndProcedure
\end{algorithmic}

Before computation of chord length function an input array must be pre-processed
to find an interface for the phase of interest. For this purpose we use
distance transform function defined as follows:
\begin{equation}
  \mathcal{D}(\bm{x})= \left\{
  \begin{array}{ll}
    0 & \quad \bm{x} \in \text{solid phase} \\
    \min\limits_{y \in \text{solid phase}} \rho(\bm{x},\bm{y}) & \quad \text{otherwise}
  \end{array}
\right. \label{eq:distance-transform}
\end{equation}
There are many algorithms for distance transform, including algorithms performed
in linear time \cite{DT}.In \code{chord\_length} function we firstly apply
distance transform to an input to find the interface and then accumulate chord
length in per-slice manner. Finally, chord lengths can be binned to a histogram.
\begin{algorithmic}[1]
  \Procedure{chord\_length}{$A, phase$}
    \State $A' \gets I^{(phase)}(A)$
    \State $A_E \gets \mathcal{D}(A') = 1$
    \Comment We consider elements for which the distance transform equals to $1$
    to be part of the interface
    \State $chord\_list \gets [\ ]$
    \ForAll{$(S', S_E) \in zip(slices(A'), slices(A_E))$}
      \Comment iterate through all one-dimensional slices in $A'$ and
      corresponding slices in $A_E$
      \State $l \gets 0$
      \State $maybe\_chord \gets False$
      \For{$idx \gets 0, length(S')-1$}
        \If{$S_E[idx]$}
          \Comment Interface found
          \If{$l > 1\ and\ maybe\_chord$}
            \Comment Chord found
            \State $chord\_list \gets l:chord\_list$
          \EndIf
          \State $l \gets 0$
          \State $maybe\_chord \gets True$
        \ElsIf{$\neg S'[idx]$}
          \Comment We are not in the phase of interest
          \State $maybe\_chord \gets False$
        \EndIf
      \EndFor
    \EndFor
    \State \textbf{return} $chord\_list$ or $histogram(chord\_list)$
  \EndProcedure
\end{algorithmic}
This implementation also works in linear time.

Another function built on top of distance transform is the pore size function
$P(r)$. Reckon that pore size function is a probability distribution
function. In practice, it is conveniently represented as a histogram of pore sizes
found in the input image. Our algorithm for calculating $P(r)$ is as follows:
\begin{algorithmic}[1]
  \Procedure{$P$}{$A$}
    \State $D \gets \mathcal{D}(A)$
    \Comment Apply $\mathcal{D}$ (\cref{eq:distance-transform}) to each element of $A$
    \State $D' = \{ x \in D: x \ne 0\}$
    \Comment Remove zeros
    \State Bin values in $D'$ to histogram $H$
    \State \textbf{return} $H$
    \Comment Alternatively just return $D'$
  \EndProcedure
\end{algorithmic}
Algorithmic efficiency is $O(n)$ where $n$ is the number of elements in the
input array.

>>>>

The interface between phases must not change rapidly for the edge filter to
work correctly. Therefore, the algorithms for surface functions require input
images to have apropriate resolution. A criterion for goodness can be expressed with
$C_{\alpha}$ parameter. Let $\alpha \in [0, 1]$ and $f(x)$ be
(band-limited) input image ($x \in \mathbb{R}^2$ or $\mathbb{R}^3$). We define
$C_\alpha$ as follows:
\begin{align*}
  f_0(x) &= f(x) - \langle f(x) \rangle \\
  C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
    dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
\end{align*}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. We have previously
shown that criterion $C_{0.5} > 0.97$ holds then the image is suitable for
robust $F_{ss}$ and $F_{sv}$ evaluation \cite{samarin2023robust}.  Note that
function \code{lowfreq\_energy\_ratio} that computes $C_{0.5}$ parameter for any
input binary image is the part of \code{CorrelationFunctions.jl} package.

@@

Cluster function can be calculated using the same technique as in map approach,
but if the number of clusters is high, a naïve $O(n^2)$ algorithm may be
preferred to $O(MC \log M)$ full map algorithm. Such an approach is present below:
\begin{algorithmic}[1]
  \Procedure{c2}{$A, phase, periodic$}
    \State $A' \gets I^{(phase)}(A)$
    \State $CC \gets label\_components(A', periodic)$
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \ForAll{$slice \in slices(CC)$}
      \For{$i = 0,len-1$}
        \State $end \gets len-1$
        \If{$periodic$}
          \State $end \gets end + i$
        \EndIf
        \For{$j = i, end$}
          \If{$slice[i] = slice[j \mod len] \ne 0$}
            \Comment $0$ is a label for background.
            \State $dst \gets |(j \mod len) - i|$
            \State $result[dst] \gets result[dst] + 1$
          \EndIf
        \EndFor
      \EndFor
    \EndFor
    \State \textbf{return} $result$ divided by number of trials
  \EndProcedure
\end{algorithmic}

Before computation of chord length function an input array must be pre-processed
to find an interface for the phase of interest. For this purpose we use
distance transform function defined as follows:
\begin{equation}
  \mathcal{D}(\bm{x})= \left\{
  \begin{array}{ll}
    0 & \quad \bm{x} \in \text{solid phase} \\
    \min\limits_{y \in \text{solid phase}} \rho(\bm{x},\bm{y}) & \quad \text{otherwise}
  \end{array}
\right. \label{eq:distance-transform}
\end{equation}
There are many algorithms for distance transform, including algorithms performed
in linear time \cite{DT}.In \code{chord\_length} function we firstly apply
distance transform to an input to find the interface and then accumulate chord
length in per-slice manner. Finally, chord lengths can be binned to a histogram.
\begin{algorithmic}[1]
  \Procedure{chord\_length}{$A, phase$}
    \State $A' \gets I^{(phase)}(A)$
    \State $A_E \gets \mathcal{D}(A') = 1$
    \Comment We consider elements for which the distance transform equals to $1$
    to be part of the interface
    \State $chord\_list \gets [\ ]$
    \ForAll{$(S', S_E) \in zip(slices(A'), slices(A_E))$}
      \Comment iterate through all one-dimensional slices in $A'$ and
      corresponding slices in $A_E$
      \State $l \gets 0$
      \State $maybe\_chord \gets False$
      \For{$idx \gets 0, length(S')-1$}
        \If{$S_E[idx]$}
          \Comment Interface found
          \If{$l > 1\ and\ maybe\_chord$}
            \Comment Chord found
            \State $chord\_list \gets l:chord\_list$
          \EndIf
          \State $l \gets 0$
          \State $maybe\_chord \gets True$
        \ElsIf{$\neg S'[idx]$}
          \Comment We are not in the phase of interest
          \State $maybe\_chord \gets False$
        \EndIf
        \State $l \gets l+1$
      \EndFor
    \EndFor
    \State \textbf{return} $chord\_list$ or $histogram(chord\_list)$
  \EndProcedure
\end{algorithmic}
This implementation also works in linear time.

Another function built on top of distance transform is the pore size function
$P(r)$. Recall that pore size function is a probability distribution
function. In practice, it is conveniently represented as a histogram of pore sizes
found in the input image. Our algorithm for calculating $P(r)$ is as follows:
\begin{algorithmic}[1]
  \Procedure{$P$}{$A$}
    \State $D \gets \mathcal{D}(A)$
    \Comment Apply $\mathcal{D}$ (\cref{eq:distance-transform}) to each element of $A$
    \State $D' = \{ x \in D: x \ne 0\}$
    \Comment Remove zeros
    \State Bin values in $D'$ to histogram $H$
    \State \textbf{return} $H$
    \Comment Alternatively just return $D'$
  \EndProcedure
\end{algorithmic}
Algorithmic efficiency is $O(n)$ where $n$ is the number of elements in the
input array.
